---
title: "Data Cleaning: Road Network + LiDAR Integration"
author: "Aarav Gosalia 42576439"
date: "2025-11-04"
output: html_document
---

# 1. Overview
This document outlines the complete data-processing pipeline used to construct an elevation-aware road network for the City of Kelowna. The workflow integrates authoritative road centerline data from the BC Data Catalogue with LiDAR-derived elevation information to produce a 3D-enhanced graph representation suitable for network analysis, slope estimation, and 3D distance computation.

The processing steps include:
1. Importing and standardizing road geometry
The raw road centerlines are downloaded via Web Feature Service (WFS) and transformed into a consistent projected coordinate system (BC Albers – EPSG:3005).

2. Cleaning and preparing road segments
Multiline geometries are normalized to simple LINESTRING objects, per-segment lengths are computed, and attributes such as road name, class, and lane count are extracted.

3. Generating graph edges
Each road segment is converted into an edge, preserving attributes and assigning unique IDs.

4. Extracting graph nodes from segment endpoints
Segment start and end coordinates are snapped, deduplicated, and indexed to form a consistent node table.

5. Constructing and visualizing the 2D road network graph
Nodes and edges are assembled into an undirected graph structure and plotted for verification.

6. Loading and mosaicking LiDAR-derived DEM tiles
Multiple DEM tiles (1 m resolution) are reprojected into BC Albers and mosaicked into a single elevation raster covering the available region.

7. Extracting elevation for network nodes
Only nodes falling inside the DEM extent receive elevation values; these form the basis for computing slope and 3D length.

8. Filtering edges to DEM-covered areas and computing slope metrics
Edges whose endpoints both lie within the DEM region are retained, joined with elevation values, and enriched with elevation difference, slope, and 3D geometric length.

## 2. Setup
```{r setup, message=FALSE, warning=FALSE}
library(sf)
library(dplyr)
library(readr)
library(tidyr)
library(tibble)
library(igraph)
library(terra)

# Paths
gpkg_path <- "data_kelowna/DRA_Kelowna_Subset.gpkg"
out_dir   <- "data_kelowna"

# Load roads
roads_raw <- st_read(gpkg_path, quiet = TRUE)

# Ensure projected CRS in meters (BC Albers)
if (st_crs(roads_raw)$epsg != 3005) {
  roads_raw <- st_transform(roads_raw, 3005)
}

# Normalize: cast to LINESTRING (explode MULTILINESTRING)
roads <- roads_raw |> 
  st_cast("LINESTRING", warn = FALSE) |>
  st_make_valid()

# Add a stable segment id
roads <- roads |> mutate(seg_id = row_number())

# Quick look
print(st_crs(roads))
nrow(roads)
```

# 3. Create Cleaned Edge Table

We extract relevant road attributes and compute two versions of length:
- FEATURE_LENGTH_M: length as stored in DRA
- Computed (st_length): length measured from geometry
```{r}
# Create two length columns: official vs computed
roads$length_feature_m <- roads$FEATURE_LENGTH_M # official DRA length
roads$length_calc_m     <- as.numeric(st_length(st_geometry(roads))) # computed length

#Clean attribute table
edges <- roads %>%
  mutate(
    road_name = coalesce(
      ROAD_NAME_FULL, ROAD_NAME_ALIAS1, ROAD_NAME_ALIAS2,
      ROAD_NAME_ALIAS3, ROAD_NAME_ALIAS4
    )
  ) %>%
  st_drop_geometry() %>%
  select(
    edge_id = DIGITAL_ROAD_ATLAS_LINE_ID,
    road_name,
    road_class   = ROAD_CLASS,
    road_surface = ROAD_SURFACE,
    lanes        = NUMBER_OF_LANES,
    highway_rt   = HIGHWAY_ROUTE_NUMBER,
    length_feature_m,
    length_calc_m
  )

write_csv(edges, file.path(out_dir, "edges.csv"))
```

# 4. Extract Road Endpoints (Nodes)

Road geometries are converted to node coordinates by extracting start and end points for each LINESTRING.

```{r}
# Extract coordinates for each vertex in each road line
coords <- st_coordinates(roads)

#Attach edge ID to each coordinate set (uses L1 indexing)
coords_df <- cbind(coords, edge_id = roads$DIGITAL_ROAD_ATLAS_LINE_ID[coords[,"L1"]])

# Compute endpoints for each road segment
endpoints <- coords_df %>%
  as_tibble() %>%
  group_by(edge_id) %>%
  summarize(
    start_x = first(X),
    start_y = first(Y),
    end_x   = last(X),
    end_y   = last(Y),
    .groups = "drop"
  )

# Convert from wide to long format
endpoints_long <- endpoints %>%
  pivot_longer(
    cols = c(start_x, start_y, end_x, end_y),
    names_to = c("pos", ".value"),
    names_pattern = "(start|end)_(x|y)"
  )

# Snap roads to grid and assign node IDs
tol <- 0.5
endpoints_long <- endpoints_long %>%
  mutate(
    x_r = round(x / tol) * tol,
    y_r = round(y / tol) * tol,
    node_key = paste0(x_r,"_",y_r)
  )

# Node table
nodes <- endpoints_long %>%
  distinct(node_key, x = x_r, y = y_r) %>%
  arrange(node_key) %>%
  mutate(node_id = row_number())

# Associate endpoints with node IDs
edge_nodes <- endpoints_long %>%
  left_join(nodes, by = c("node_key","x_r" = "x","y_r" = "y")) %>%
  select(edge_id, pos, node_id) %>%
  pivot_wider(names_from = pos, values_from = node_id) %>%
  rename(from = start, to = end)

# Final edge table with from/to
edges_with_nodes <- edges %>%
  left_join(edge_nodes, by = "edge_id") %>%
  filter(!is.na(from), !is.na(to))

# Save outputs
write_csv(nodes %>% select(node_id, x, y), file.path(out_dir, "nodes.csv"))
write_csv(edges_with_nodes, file.path(out_dir, "edges_with_nodes.csv"))
```

# 5. Vizualize the Road Network and Graph

```{r}
nodes <- read_csv("data_kelowna/nodes.csv")
edges_with_nodes <- read_csv("data_kelowna/edges_with_nodes.csv")
roads <- st_read("data_kelowna/DRA_Kelowna_Subset.gpkg", quiet = TRUE)

nodes_sf <- st_as_sf(nodes, coords = c("x", "y"), crs = st_crs(roads))

# Plot Nodes + Roads
plot(st_geometry(roads), col = "black", lwd = 0.6, main = "Kelowna Road Network")
plot(st_geometry(nodes_sf), col = "maroon", pch = 20, cex = 0.3, add = TRUE)

# Remove self-edges
edges_clean <- edges_with_nodes %>% 
  filter(from != to)

# Build igraph object
g <- graph_from_data_frame(
  d = edges_clean %>% select(from, to),
  vertices = nodes %>% mutate(node = node_id),
  directed = FALSE
)

# Plot
plot(
  g,
  vertex.size = 2,
  vertex.color = "blue",
  vertex.frame.color = NA,
  vertex.label = NA,
  edge.color = "black",
  edge.width = 0.3,
  edge.arrow.size = 0,
  main = "Kelowna Road Network (Clean Graph View)"
)
```

# 6. Load and Prepare DEM Tiles (LiDAR-derived elevation)

This section loads all LiDAR DEM tiles, reprojects them to EPSG:3005, verifies the CRS of each, and mosaics (joins) tiles.

```{r}
dem_files <- list.files("data_kelowna/tiles/", pattern="\\.tif$", full.names=TRUE)

# Load + Reproject each tile to BC Albers (EPSG:3005)
dem_list <- lapply(dem_files, function(f) {
  r <- rast(f)
  project(r, "EPSG:3005", method="bilinear")
})

# Check CRS of all DEM tiles
sapply(dem_list, crs)

# Mosaic Tiles (Join)
dem_3005 <- Reduce(function(x, y) mosaic(x, y, fun=mean), dem_list)
```

# 7. Extract Elevation for Road-Network Nodes
This section integrates the road network and DEM by assigning elevation to each node inside DEM extent.

```{r}
# Check Spatial Overlap: Road Network vs DEM Coverage
# We ensure extraction is valid and only process nodes inside DEM coverage.

roads_bbox <- st_bbox(roads_raw)
dem_bbox   <- as.vector(ext(dem_3005))

roads_bbox
dem_bbox

plot(st_geometry(roads_raw), col="black", main="Roads vs DEM extent")
plot(ext(dem_3005), add=TRUE, border="red", lwd=3)

# Convert nodes to sf
nodes_sf <- st_as_sf(nodes, coords = c("x", "y"), crs = st_crs(roads))

# Detect which nodes fall inside the DEM
xy <- st_coordinates(nodes_sf)
cells <- terra::cellFromXY(dem_3005, xy)
inside_idx <- which(!is.na(cells))

#Extract node elevation
nodes_inside <- nodes_sf[inside_idx, ]

elev_vals <- terra::extract(dem_3005, vect(nodes_inside))[[1]]
nodes_inside$elev_m <- elev_vals

#Save final nodes csv
write_csv(nodes_inside, "data_kelowna/final_nodes.csv")
```

# 8. Subset the Road Graph to DEM-covered Edges + Compute Slope

```{r}
valid_nodes <- nodes_inside$node_id

# Filter edges to keep only those whose endpoints BOTH lie inside the DEM area
edges_inside <- edges_with_nodes %>%
  filter(from %in% valid_nodes & to %in% valid_nodes)

# Join elevation values from nodes to edges
edges_inside <- edges_inside %>%
  # Add elevation of the start node
  left_join(nodes_inside %>% select(node_id, elev_m),
            by = c("from" = "node_id")) %>%
  rename(elev_from = elev_m) %>%
  
  # Add elevation of the end node
  left_join(nodes_inside %>% select(node_id, elev_m),
            by = c("to" = "node_id")) %>%
  rename(elev_to = elev_m) %>%
  
  # Compute slope and 3D geometric length
  mutate(
    # Elevation change between endpoints (positive = uphill)
    elev_diff = elev_to - elev_from,
    
    # Slope = rise/run (unitless)
    slope = elev_diff / length_calc_m,
    
    # 3D geometric length:
    # Square root of (horizontal_length^2 + elevation_change^2)
    length_3d_m = sqrt(length_calc_m^2 + elev_diff^2)
  )

# Save final edges csv
write_csv(edges_inside, "data_kelowna/final_edges.csv")

# Visualize slope distribution
hist(
  edges_inside$slope,
  breaks = 50,
  col = "orange",
  border = "white",
  main = "Slope Distribution of Road Segments with DEM Coverage",
  xlab = "Slope (rise/run)"
)

```